Spatial variation of the native colon microbiota in healthy adults

### 2017-10-31 submission/manuscript.tex
updated manuscript EOD tuesday

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={The microbiota of the proximal and distal human colon},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{The microbiota of the proximal and distal human colon}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

\vspace{35mm}

Running title: The microbiome of the proximal and distal human colon

\vspace{35mm}

Kaitlin J. Flynn\textsuperscript{1}, Charles C.
Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2},
Danielle Kimberly Turgeon\textsuperscript{3\(\dagger\)}, and Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

\vspace{35mm}

\(\dagger\) Corresponding authors:
\href{mailto:kturgeon@med.umich.edu}{\nolinkurl{kturgeon@med.umich.edu}}
and \href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

@@

2. Pennslyvania State University, Hershey, Pennslyvania ??

@@

Chemical and nutrient gradients along the human colon create
microenvironments that affect the distrubution and composition of the
gut microbiota. The microbiome has been implicated in colorectal cancer
(CRC) and inflammatory bowel disease (IBD). Further, these diseases
exhibit different symptoms depending on the location of the colon they
are found in. CRC tumors of the proximal and distal colon are
morphologically and genetically distinct. Similarly, inflammatory bowel
diseases such as Crohn's are typically exacerbated in the proximal
instestine while ulcerative colitis patients often experience symptoms
in the distal colon. Previous analysis of the fecal microbiota from
healthy and CRC or IBD patients has revealed different microbial
signatures associated with these diseases. We extended these
observations of the fecal microbiome to include analysis of the proximal
and distal healthy human colon. We used a two-colonoscope approach on
subjects that had not undergone standard bowel preparation procedure. In
contrast to previous efforts using prepared colonoscopy, this technique
allowed us to characterize the native proximal and distal luminal and
mucosal microbiome without prior chemical disruption. 16S rRNA gene
sequencing was performed on proximal and distal mucosal and luminal
biopsies and home-collected stool for 20 healthy individuals. Diversity
analysis revealed that each site contained a diverse community and that
a patient's samples were more similar to each other than to that of
other individuals. A patient's feces were most similar to samples taken
from the distal lumen, likely reflecting the anatomical structure of the
colon. Since we could not differentiate the communities in the proximal
and distal colon based on community structure or community membership
alone, we employed the Random Forest machine-learning algorithm to
identify key taxa that the two sites in the lumen and mucosa. Random
Forest classification models were built using taxa abundance and sample
location and revealed distinct populations that were found in each
location. A model differentiating the proximal mucosa and lumen was
built with an AUC of 0.856. The proximal mucosa had a higher abundance
of the genera \emph{Enterobacteriaceae} and \emph{Bacteriodes}.
\emph{Peptoniphilus, Anaerococcus, Finegoldia,} and \emph{Turicibacter}
were most likely to be found in distal mucosal samples versus distal
luminal samples (AUC = 0.975). The classification model performed well
(AUC = 0.920) when classifying mucosal samples into proximal or distal
sides, but separating luminal samples from each side proved more
challenging (AUC = 0.575). The distal mucosa was found to have high
populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
Proximal and distal luminal samples were comprised of many of the same
taxa, likely reflecting the fact that stool moves along the colon from
the proximal to distal end. By sampling the unprepped human colon, our
results have identified distinct bacterial populations native to the
proximal and distal sides. Further investigation of these bacteria may
elucidate if and how these groups contribute to different pathogenesis
processes on the respective sides of the colon.

\paragraph{wordcount?}\label{wordcount}

@@

The human colon is an ecosystem comprised of several different
microenvironments inhabited by resident bacterial members of the
microbiome. Concentrations of oxygen, water and anti-microbial peptides
change along the gut axis and influence what populations of microbes
reside in each location. Microenvironments differ not only
longitutinally along the colon, but latitudinally from the epithelium to
mucosa to intestinal lumen, offering several sites for different
microbial communities to flourish. The identity of these specific
microbes and communities are important for understanding the etiology of
complex colon diseases such as Inflammatory Bowel Disease (IBD) and
Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
accelerated by perturbations in gut microbes (1--3). The severity,
symptoms, morbidity and mortality of these diseases is known to vary
based upon the biogeographical location in which they occur. For
instance, CRC tumors that arise on the distal side of the colon are
infiltrating lesions that present with painful symptoms (4). In
contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
are sessile and form along the wall of the colon, often remaining
asymptomatic until advanced carcinogenesis (4). The distal and proximal
sides of the colon differ in the amount of inflammation present and the
genomic instability of precancerous cells, respectively, in addition to
variation in the previously mentioned chemical gradients (1, 5, 6). In
IBD patients, disease flares in the distal colon are usually indicative
of ulcerative colitis (UC), whereas Crohn's disease (CD) patients
typically experience disease in the small intestine, ileum and proximal
colon (2). UC presents as large and highly inflammed mucosal ulcers,
where as CD lesions are often smaller and have areas of normal tissue
distributed amongst the flare (2). Thus, given the varied physiology of
the proximal-distal axis of the colon and known differences in disease
patterns at these sites, symbiotic microbes and their metabolites likely
vary as well, and may influence the heterogenous disease prognoses of
IBD and CRC. Because CRC can be a long-term complication of IBD, the
distribution of microbes is important to understanding the
pathophysiology of both diseases.

Several recent findings have shown that development and progression of
IBD or CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host (1, 3). For
instance, comparison of the bacteria present on CRC tumors with those
found on nearby healthy tissue has identified specific species that are
tumor-associated (7). These species include the oral pathogens
\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
Interestingly, these periodontal pathogens have been highly predictive
of whether a patient had CRC tumors or not in our prior human stool
classification studies (8). \emph{F. nucleatum} has also been found to
be elevated in the stool and biopsies of patients with IBD as compared
to healthy controls (9). Furthermore, studies of \emph{F. nucleatum}
isolated from mucosal biopsies showed that more invasive \emph{F.
nucleatum} positively correlates with IBD disease level (9). Like many
intestinal pathogens, the bacteria appear to have a high-impact despite
being lowly-abundant in the community (2). The physiology of these rare
taxa may contribute to the colonic disease state. These studies often
examined only shed human stool or the small intestine, preventing
fine-resolution analysis of paired samples from the proximal and distal
sides of the colon. Similarly, comparisons of on- or off-tumor/lesion
bacteria rarely have matched tissue from the other side of the colon
from the same patient, limiting what conclusions can be drawn about the
colonic microbiome overall, let alone at that specific site. Due to
these limitations, the contribution of the gut microbiota to IBD and CRC
disease location in the colon is largely undefined. Characterizing these
communities could provide needed insight into disease etiology,
including how the disruption of the healthy community could promote the
initiation or proliferation of the distinct proximal and distal CRC
tumors or IBD flares.

The few existing profiles of the microbial biogeography of the gut have
been limited by sample collection methods. The majority of human gut
microbiome studies have been performed on whole shed feces or on samples
collected during colonoscopy procedures. While the latter method allows
investigators to acquire samples from inside the human colon, typically
this procedure is preceded by the use of bowel preparation methods such
as the consumption of laxatives to cleanse the bowel. Bowel preparation
is essential for detecting cancerous or precancerous lesions in the
colon, but complicates microbiome profiling as the chemicals strip the
bowel of contents and disrupt the mucosal layer (10, 11). As such, what
little information we do have about the biogeographical distribution of
the microbes in the proximal and distal colon is confounded by the bowel
preparation procedure.

Here we address the limitations of previous studies and effectively
identify the microbes specific to the lumen and mucosa of the proximal
and distal healthy human colon. We used an unprepared colonoscopy
technique to sample the natural community of each location of the gut
without prior disruption of the native bacteria in 20 healthy
volunteers. To address the inherent inter-individual variation in human
microbiomes, we used a machine-learning classification algorithm trained
on curated 16S rRNA sequencing reads to identify microbes specific to
each location. We found that our classification models were able to
separate mucosal and lumenal samples as well as differentiate between
sides of the colon based on populations of specific microbes. By
identifying the specific microbes we are poised to ask if and how the
presence or disruption of the microbes at each site contribute to the
development of the specific tumor subtypes of CRC in the proximal and
distal human colon.

@@

Lumenal and mucosal samples were collected from the proximal and distal
colon of 20 healthy humans that had not undergone bowel preparation
(Figure 1). Participants also collected stool at home one week prior to
the procedure. To characterize the bacterial communities present at
these sites, 16S rRNA gene sequencing was performed on extracted DNA
from each sample. As expected, each site was primarily dominated by
\emph{Firmicutes} and \emph{Bacteriodetes} (Figure 2A) (12). Samples had
varying levels of diversity at each site, irrespective of the individual
(Figure 2B). For example, the proximal mucosa was more diverse than the
distal for some individuals while the opposite was true for others.
Therefore we could not identify a clear pattern of changes in microbial
diversity along the gut axis.

To compare similarity between sides (proximal or distal) or sites (lumen
or mucosa), we calculated distances from Operational Taxonomic Unit
(OTU) abundances and compared these distances for all individuals.
Again, across all patient samples we observed a range of distances when
comparing sample locations (Figure 3A) and again those ranges did not
follow a clear pattern on an individual basis. However, when comparing
median distances between the proximal lumen and mucosa, the proximal
versus distal lumen, the proximal versus distal mucosa, and the distal
lumen and mucosa, we found that the proximal lumen and mucosa were most
similar to each other than the other samples (P \textless{} 0.005,
Wilcoxon, BH adjustment).

\paragraph{Fecal samples resemble lumenal samples from the distal
colon}\label{fecal-samples-resemble-lumenal-samples-from-the-distal-colon}

Next, we calculated distances to examine how each sample compared to the
home-collected feces. Amidst variability between patients, we did
identify significantly smaller distance between the distal lumenal
sample and the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
adjustment). Furthermore, there was an even larger difference in the
comparisons of the distal mucosa to the feces, indicating that the
mucosa is different from the stool as compared to lumen (P \textless{}
0.0005, Wilcoxon, BH adjustment). To determine what factors may be
driving the differences seen among the samples, we compared distances
between samples from all subjects (interpersonal) versus samples from
within one subject (intrapersonal). We found that samples from one
individual were far more similar to each other than to other study
subjects (Figure 3C), consistent with previous human microbiome studies
that have sampled multiple sites of the human colon (13--15). Thus
interpersonal variation between subjects drove the differences between
samples more than sample site or location. Overall, the results
comparing the structure of the communities suggest that the contents of
the distal lumen are most representative of the patient's feces, and the
microbes remaining on the mucosa are more distinct.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models trained using OTU abundances.
We used 10-fold cross validation to build the first model to classify
the lumenal versus mucosal samples for the proximal and distal sides,
independently (Figure S1A). The models were constructed to only use the
five most predictive OTUs as input to reduce overfitting between
samples. The models performed well when classifying these samples (AUC =
0.856 and AUC = 0.975, respectively). The OTUs that were most predictive
of each site were identified by their greatest mean decrease in accuracy
when removed from the model. For distinguishing the proximal lumen and
mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
were differentially abundant (Figure 4A). The model classifying the
distal lumen and mucosa identified OTUs from \emph{Turicibacter},
\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
\emph{Anaerococcus} genera (Figure 4B). These results indicated that
there were fine differences between the different sites of the colon,
and that these could be traced to specific OTUs on each side.

Next, we built a model to differentiate the proximal and distal lumenal
samples using 10-fold cross validation. The model performed best when
distinguishing the proximal versus distal mucosa (Figure S1B, AUC =
0.920) compared to the proximal versus distal lumen (AUC = 0.575). These
models were also constructed to use only the five most predictive OTUs
as input to reduce overfitting. OTUs that were differentially abundant
between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5A).
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5B). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

\paragraph{Bacterial OTUs associated with cancer are found in healthy
individuals}\label{bacterial-otus-associated-with-cancer-are-found-in-healthy-individuals}

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors.
(16). There were 14 patient samples with the \emph{F. nucleatum subsp.
animalis} sequences. Of the samples with the highest abundance of
\emph{F. nucleatum subsp. animalis}, four of the samples were from the
proximal mucosa and three from the distal mucosa (Figure 6A). The second
most frequent \emph{Fusobacterium} sequence was OTU472, which aligned
with 99\% identity to \emph{F. varium}. In addition to \emph{F.
nucleatum}, \emph{F. varium} has been associated with IBD (17). Four
study participants harbored \emph{F. varium} and the samples were split
evenly between the proximal and distal mucosa (Figure 6B). OTU152 was
similar to the members of the \emph{Porphyromonas} genus and the most
frequent sequence in that OTU aligned to \emph{Porphyromonas
asacharolytica} (99\% over full length), another bacterium commonly
detected and isolated from colorectal tumors. OTU152 was only detected
on the distal mucosa, and in fact was one of the OTUs the classification
model identified as separating distal and proximal sides (Figure 6C).
Among the 11 distal mucosa samples that were positive for \emph{P.
asacharolytica}, the relative abundances for this OTU ranged from 0.01\%
- 16\%. Thus, disease-associated OTUs could be found in our sample set
of 20 healthy individuals.

@@

Here we identified bacterial taxa that were specific to the lumen and
mucosa of the proximal and distal sides of the human colon from samples
collected during unprepared colonoscopy. We found that all locations
contained a range of phyla abundances and a range of diversity, but that
there was a wide variability between subjects. Pairwise comparisons of
each of the sites revealed that the proximal mucosa and lumen were most
similar to each other. Further, comparison of colonoscopy-collected
samples with samples collected from stool at home showed that the distal
lumen was most similar to feces. Random Forest models built on OTU
relative abundances from each sample identified microbes that were
particular to each location of the colon. Finally, we were able to
detect some bacterial OTUs associated with colonic disease in our
healthy patient cohort. Using unprepped colonoscopy and machine
learning, we have identified bacterial taxa specific to the healthy
proximal and distal human colon.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Spatial variation of the native colon microbiota in healthy adults},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Spatial variation of the native colon microbiota in healthy adults}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

\vspace{15mm}

Running title: Spatial variation of native colon microbiota

\vspace{15mm}

Kaitlin J. Flynn\textsuperscript{1}, Charles C.
Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2}, D.
Kim Turgeon\textsuperscript{3\(\dagger\)}, and Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

\vspace{15mm}

\(\dagger\) Corresponding authors:
\href{mailto:kturgeon@umich.edu}{\nolinkurl{kturgeon@umich.edu}} and
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

@@

2. Pennsylvania State University, Hershey, Pennsylvania ??

@@

\textbf{Background and Aims:} The microbiome has been implicated in the
development of colorectal cancer (CRC) and inflammatory bowel diseases
(IBD). The specific traits of these diseases vary along the axis of the
digestive tract. Further, variation in the structure of the gut
microbiota has been associated with both diseases. Here we profiled the
microbiota of the healthy proximal and distal mucosa and lumen to better
understand how bacterial populations vary along the colon.

\textbf{Methods:} We used a two-colonoscope approach to sample proximal
and distal mucosal and luminal contents from the colons of 20 healthy
subjects that had not undergone any bowel preparation procedure. The
biopsies and home-collected stool were subjected to 16S rRNA gene
sequencing and Random Forest classification models were built using taxa
abundance and location to identify microbiota specific to each site.

\textbf{Results:} : The right mucosa and lumen had the most similar
community structures of the five sites we considered from each subject.
The distal mucosa had higher relative abundance of Finegoldia,
Murdochiella, Peptoniphilus, Porphyromonas and Anaerococcus. The
proximal mucosa had more of the genera Enterobacteriaceae, Bacteroides
and Pseudomonas. The classification model performed well when
classifying mucosal samples into proximal or distal sides (AUC=0.850).
Separating proximal and distal luminal samples proved more challenging
(AUC=0.580) and specific microbiota that differentiated the two were
hard to identify.

\textbf{Conclusions:} By sampling the unprepped colon, we identified
distinct bacterial populations native to the proximal and distal sides.
Further investigation of these bacteria may elucidate if and how these
groups contribute to different disease processes on their respective
sides of the colon.

Words: 259/260

\textbf{Keywords}: microbiome; colon cancer; proximal and distal colon

@@

The human colon is an ecosystem comprised of numerous microenvironments
that select for different microbiota. Concentrations of oxygen, water
and anti-microbial peptides change along the gut axis and influence
which microbiota reside in each location. Microenvironments differ not
only longitudinally along the colon, but also radially from the
epithelium to mucosa to intestinal lumen, offering several sites for
different microbial communities to flourish. The identity of these
specific microbiota and communities are important for understanding the
etiology of complex colon diseases such as Colorectal Cancer (CRC) and
Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or
accelerated by perturbations the structure of the gut microbiota (1--3).
The manifestations of these diseases are known to vary based upon the
location in which they occur. For instance, CRC that arises in the
distal (left) colon are of hindgut origin and tend to have large
chromosomal alterations indicative of chromosomal instability (1). In
contrast, CRC arising in the proximal (right) colon tumors are of midgut
origin and tend to be sessile and microsatellite instable (MSI with BRAF
and KRAS mutations) (1). In addition to the environmental gradients
within the colon, the distal and proximal sides of the colon differ in
the amount of inflammation present and the genomic instability of
precancerous cells, respectively (1,4,5). In IBD patients, disease
occurring in the distal colon extending proximally is usually indicative
of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur
anywhere along the GI tract, most commonly in the ileum and the cecum
(2). UC presents as continuous disease with only mucosal involvement,
where as CD has skip lesions and full thickness involvement that may
cause abscesses, strictures and fistulas (2). Thus, given the varied
physiology of the proximal-distal axis of the colon and known
differences in disease patterns at these sites, symbiotic microbiota and
their metabolites likely vary as well, and may influence the
heterogeneous disease prognoses of IBD and CRC. Because CRC can be a
long-term complication of IBD, the distribution of microbiota is
important to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of
IBD or CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host (1,3,6). For
instance, comparison of the bacteria present on CRC tumors with those
found on nearby healthy tissue has identified specific species that are
tumor-associated (7). Specific bacteria have also been identified in
fecal samples of patients with varying stages of colon tumorigenesis
(8,9). These species include the oral pathogens \emph{Fusobacterium
nucleatum} and \emph{Porphyromonas asacharolytica}. \emph{F. nucleatum}
has also been found to be elevated in the stool and biopsies of patients
with IBD as compared to healthy controls (10,11). Furthermore, studies
of \emph{F. nucleatum} isolated from mucosal biopsies showed that more
invasive \emph{F. nucleatum} positively correlates with IBD disease
level (10). Like many intestinal pathogens, the bacteria appear to have
a high-impact despite being lowly-abundant in the community (2). The
physiology of these rare taxa may contribute to the colonic disease
state. These studies often examined only shed human stool or the small
intestine, preventing fine-resolution analysis of paired samples from
the proximal and distal sides of the colon. Similarly, comparisons of
on- or off-tumor/lesion bacteria rarely have matched tissue from the
other side of the colon from the same, disease-baring patient, limiting
what conclusions can be drawn about the colonic microbiome overall, let
alone at that specific site. Due to these limitations, the contribution
of the gut microbiota to CRC and IBD disease location in the colon is
largely undefined. Characterizing these communities in healthy
individuals could provide needed insight into disease etiology,
including how the disruption of the healthy community could promote the
initiation or proliferation of the distinct proximal and distal CRC
tumors or IBD flares.

The few existing profiles of the microbial spatial variation of the
colon have been limited by sample collection methods. The majority of
human gut microbiome studies have been performed on whole shed feces or
on samples collected during colonoscopy or surgery. While the latter
method allows investigators to acquire samples from inside the human
colon, typically these procedures are preceded by the use of bowel
preparation methods such as the consumption of laxatives to cleanse the
bowel. Bowel preparation is essential for detecting cancerous or
precancerous lesions in the colon, but complicates microbiome profiling
as the chemicals strip the bowel of contents and disrupt the mucosal
layer (12,13). As such, what little information we do have about the
spatial distribution of the microbiota in the proximal and distal colon
is confounded by the bowel preparation procedure.

Here we address the limitations of previous studies and identify the
microbes specific to the lumen and mucosa of the proximal and distal
healthy human colon. We used an unprepared colonoscopy technique to
sample the natural community of each location of the gut without prior
disruption of the native bacteria in 20 healthy volunteers. To address
the inherent inter-individual variation in microbiota, we used a
machine-learning classification algorithm trained on curated 16S rRNA
sequencing reads to identify the microbiota that were specific to each
location. We found that our classification models were able to separate
mucosal and luminal samples as well as differentiate between sides of
the colon based on populations of particular microbiota. By identifying
the distinguishing microbiota we are poised to ask if and how the
presence or disruption of the microbiota at each site contribute to the
development of the tumor subtypes of CRC in the proximal and distal
human colon.

@@

Luminal and mucosal samples were collected from the proximal and distal
colon of 20 healthy humans that had not undergone bowel preparation
(Figure 1). Subjects also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on DNA extracted from each
sample. As expected, each site was primarily dominated by
\emph{Firmicutes} and \emph{Bacteriodetes} (Figure 2A) (14). Samples had
varying levels of diversity at each site, irrespective of the individual
(Figure 2B). For example, the proximal mucosa was more diverse than the
distal for some individuals while the opposite was true for others.
Therefore we could not identify a clear pattern of changes in microbial
diversity along the gut axis.

To compare similarity between the proximal and distal sides and within
the lumen and mucosa, we compared the community structure of these sites
based on the relative abundances of individual Operational Taxonomic
Units (OTUs). Across all subjects we observed wide variation when
comparing sample locations (Figure 3A). Those ranges did not follow a
clear pattern on an individual basis. However, when comparing median
dissimilarity between the communities found in the proximal lumen and
mucosa, the proximal and distal lumen, the proximal and distal mucosa,
and the distal lumen and mucosa, we found that the proximal lumen and
mucosa were most similar to each other than to the other samples (P
\textless{} 0.005, Wilcoxon, BH adjustment).

\paragraph{Fecal samples resemble luminal samples from the distal
colon}\label{fecal-samples-resemble-luminal-samples-from-the-distal-colon}

Next, we compared the luminal and mucosal samples to the fecal sample of
each subject. Amidst the large inter-subject variation, we did identify
significantly less dissimilarity between the distal luminal sample and
the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH adjustment).
Furthermore, there was an even larger difference in the communities
found in the distal mucosa compared to the fecal communities, indicating
that the mucosa is as different from the stool as compared to lumen (P
\textless{} 0.0005, Wilcoxon, BH adjustment). These results suggest that
the contents of the distal lumen were most representative of the
subjects' feces, and the mucosal microbiota are distinct from the fecal
and luminal communities.

\paragraph{Interpersonal community variation is greater than the
variation between
sites}\label{interpersonal-community-variation-is-greater-than-the-variation-between-sites}

To determine what factors may have driven the differences seen among the
samples, we compared the community dissimilarity between samples from
all subjects (interpersonal) versus samples from within one subject
(intrapersonal). We found that samples from one individual were far more
similar to each other than to matched samples from the other subjects
(Figure 3C); this is consistent with previous human microbiome studies
that have sampled multiple sites of the human colon (15--17). Thus
interpersonal variation drove the differences between samples more than
whether the sample came from the proximal or distal side of the colon or
from the lumen or mucosa.

@@

To identify OTUs that were distinct at each site, we constructed several
Random Forest models trained using OTU relative abundances. We used
10-fold cross validation to build the first model to classify the
luminal versus mucosal samples for the proximal and distal sides,
independently (Figure S1A). The models performed well when classifying
these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that
were most predictive of each site were identified by their greatest mean
decrease in accuracy when removed from the model. For distinguishing the
proximal lumen and mucosa, OTUs affiliated with the \emph{Bacteriodes},
\emph{Actinomyces}, \emph{Psuedomonas} and \emph{Enterobacteraceae} were
included in the best model (Figure 4A). The model to differentiate
between the distal lumen and mucosa included OTUs affiliated with the
\emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
\emph{Anaerococcus} (Figure 4B). These results indicated that there were
fine differences between the different sites of the colon, and that
these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and
distal luminal samples using 10-fold cross validation. The model
performed best when distinguishing the proximal versus distal mucosa
(Figure S1B, AUC = 0.850) whereas the model to differentiate between the
proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs
included in the model differentiating the distal and proximal mucosa
included members of the \emph{Porphyromonas}, \emph{Murdochiella},
\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} (Figure
5A). The model that attempted to separate the the proximal and distal
lumen included OTUs affiliated with the \emph{Bacteroides},
\emph{Clostridium IV} and \emph{Oscillibacter} (Figure 5B).
Interestingly, \emph{Anaerococcus} and \emph{Finegoldia} were distinct
between the mucosa and lumen and also helped to differentiate between
the proximal and distal sides.

\paragraph{Bacterial OTUs associated with CRC and IBD are found in
healthy
individuals}\label{bacterial-otus-associated-with-crc-and-ibd-are-found-in-healthy-individuals}

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligned via blastn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors
(18). There were 14 samples from 8 subjects with the \emph{F. nucleatum
subsp. animalis} sequences. Of the samples with the highest relative
abundance of \emph{F. nucleatum subsp. animalis}, four of the samples
were from the proximal mucosa and three from the distal mucosa (Figure
S2A). The second most frequent \emph{Fusobacterium} sequence was OTU472,
which aligned with 99\% identity to \emph{F. varium}. In addition to
\emph{F. nucleatum}, \emph{F. varium} has been associated with IBD (19).
Four subjects harbored \emph{F. varium} and the samples were split
evenly between the proximal and distal mucosa (Figure S2B). OTU152 was
similar to the members of the \emph{Porphyromonas} genus and the most
frequent sequence in that OTU aligned to \emph{Porphyromonas
asacharolytica} (99\% over full length), another bacterium commonly
detected and isolated from colorectal tumors. OTU152 was only detected
on the distal mucosa, and in fact was one of the OTUs the classification
model identified as separating distal and proximal sides (Figure S2C).
Among the 11 distal mucosa samples that were positive for \emph{P.
asacharolytica}, the relative abundances for this OTU ranged from 0.01\%
to 16\%. Thus, disease-associated OTUs could be found in our sample set
of 20 healthy individuals.

@@

Here we identified bacterial taxa that were specific to the lumen and
mucosa of the proximal and distal sides of the human colon using samples
collected during an unprepared colonoscopy of healthy subjects. We found
that all locations contained a range of phyla relative abundances and a
range of diversity, but that there was a wide variability between
subjects. Pairwise comparisons of each of the sites revealed that the
proximal mucosa and lumen were most similar to each other. Further,
comparison of colonoscopy-collected samples with fecal samples
demonstrated that the distal lumen was most similar to feces. Random
Forest models built using OTU relative abundances from each sample
identified microbiota that were particular to each location of the
colon. Finally, we were able to detect some bacterial OTUs associated
with colonic disease in our healthy cohort. Using unprepped
colonoscopies and machine learning, we have identified bacterial taxa
specific to the healthy proximal and distal human colon.
